---
title: "Application"
author:
  - name: Yuxuan Wang
    affiliation: Sichuan University
    email: wyx1743@gmail.com
    orcid: "0009-0003-7817-748X"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    theme: readable
    highlight: tango
  pdf_document:
    latex_engine: xelatex
    highlight: arrow
    number_sections: false
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Application}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,eval = F
)
```

# Applications: from VasGBIF to *letsR*

This vignette demonstrates how to take the output of VasGBIF and feed it into **letsR** for macroecological analysis: constructing presence-absence matrices, mapping species richness, and integrating environmental variables.

## Preparation: running the VasGBIF pipeline

Before downstream analysis, we need a set of high-quality, non-redundant occurrence records with native-status annotations. The VasGBIF pipeline produces exactly that in five steps:

1.  **Import** — load the GBIF occurrence download ZIP
2.  **Resolve taxonomy** — match names against WCVP via TNRS
3.  **Build collection-event keys** — group records by taxon + date + coordinates
4.  **Select digital vouchers** — pick the highest-quality record per group
5.  **Refine** — validate coordinates, restore metadata, and annotate native status

Here we use the *Saxifraga records* (see <https://doi.org/10.15468/dl.4ty3ap>) as example. Substitute your own GBIF ZIP for real analyses (see the **GetRecords** vignette for download instructions).

```{r get_records}
library(rgbif)
library(dplyr)
## Records can be downloaded with rgbif by DOI without requiring authentication.
DOI <- "10.15468/dl.4ty3ap"
gbif_file <- occ_download_doi(DOI)$key %>% occ_download_get(path = tempdir()) %>% as.character()
```

```{r preparation, message=FALSE}
library(VasGBIF)
library(data.table)

# Step 1: Import 
occ_import <- import_records(path = gbif_file)

# Step 2: Parse GBIF issue flags
gbif_issue <- extract_gbif_issues(occ_import)

# Step 3: Resolve taxon names against WCVP
taxa_checked <- check_taxon(occ_import = occ_import, accuracy = 0.85)

# Step 4: Filter records by quality rules
filtered <- custom_filter(
  occ_import = occ_import,
  taxa_checked = taxa_checked,
  gbif_issue = gbif_issue
)

# Step 5: Validate coordinates and annotate native status
refined_coordinates <- refine_coordinates(
  custom_filtered = filtered,
  threads = 4
)
native_detected <- detect_native_status(
  refined_coordinates = refined_coordinates
)
```

Later we will use `refined_coordinates` and `native_detected`.

## Introducing *letsR* and Presence-Absence Matrices

***letsR*** (Vilela and Villalobos 2015) is an R package designed for macroecological analyses based on species geographic distributions. The package transforms species range maps into presence-absence matrices (PAMs), where rows represent grid cells and columns represent species. It provides a comprehensive workflow for spatial biodiversity analyses, including tools to:

1.  construct PAMs from shapefiles or occurrence points;

2.  integrate environmental variables into species distribution data;

3.  calculate species-level macroecological metrics such as range size, range overlap, and distributional midpoints;

4.  map species richness across geographic, environmental, and trait spaces;

5.  perform grid-level analyses of biodiversity patterns.

The package implements an S3 class structure (**`PresenceAbsence`**) that facilitates reproducible spatial analyses and supports integration with IUCN Red List and BirdLife data. All spatial operations are optimized for large-scale datasets, making ***letsR*** particularly suitable for comparative biogeography and conservation prioritization studies.

## Introduce Presence-Absence-Matrix

**A Presence-Absence Matrix (PAM) is the fundamental data structure in macroecological research that represents species distributions in a discrete spatial framework.**

In essence, a PAM is a two-dimensional matrix where:

- **Rows** represent discrete spatial units (grid cells) covering a geographic domain

- **Columns** represent species

- **Values** are binary: **`1`** indicates presence, **`0`** indicates absence

The first two columns typically store the longitude (x) and latitude (y) coordinates of each grid cell's centroid, allowing the matrix to be spatially referenced.

**Why PAMs matter:**

1.  **Standardization** — Converting continuous range maps (shapefiles) or scattered occurrence points into a standardized grid allows meaningful comparisons across species with vastly different distributions

2.  **Computational efficiency** — Binary matrices enable rapid calculations of biodiversity metrics (richness, turnover, endemism) across thousands of species and millions of cells

3.  **Multi-scale integration** — PAMs serve as the bridge between species-level data (traits, phylogeny, conservation status) and environmental data (climate, land use, topography)

4.  **Reproducibility** — Explicitly defined grid systems eliminate ambiguities in spatial resolution, projection, and extent

**In letsR's implementation:**

The package wraps PAMs in a **`PresenceAbsence`** S3 object that bundles:

- The presence-absence matrix itself

- A raster layer of species richness

- Species names and metadata

- Grid system parameters (resolution, projection, extent)

This design allows you to apply generic R functions (**`plot`**, **`summary`**, **`print`**) directly to the object while maintaining spatial integrity throughout analytical workflows.

## From VasGBIF Results to PAM

Filter the refined records to native occurrences, extract the columns needed by *letsR*, and construct the matrix:

```{r PAM, message=FALSE}
library(letsR)
library(stringi)

# prepare input for letsR
PAM_pre <- refined_coordinates$CoordinateCleaned[gbifID %chin% native_detected[native_status=='native',gbifID],.(
  Accepted_name,
  genus  = stri_extract_first_regex(Accepted_name, "^[^ ]+"),
  family,
  decimalLongitude,
  decimalLatitude
)]

head(PAM_pre, 5)

# Optionally filter to a specific genus or family:
# PAM_pre <- PAM_pre[genus == "Saxifraga", ]
```

Construct the Presence-Absence Matrix at 1° resolution:

```{r lets.presab.points}
PAM <- lets.presab.points(
  xy      = as.matrix(PAM_pre[, .(decimalLongitude, decimalLatitude)]),
  species = PAM_pre$Accepted_name,
  resol   = 1,     # 1-degree grid cells
  count   = FALSE  # set TRUE for progress bar on large datasets
)
```

## Summary Statistics

Basic information about the Presence-Absence object:

```{r Statistics}
summary(PAM)
names(PAM)
PAM$Presence_and_Absence_Matrix[1:10, 1:10] |>
  as.data.table()
head(PAM$Species_name)
```

Which species has the widest distribution range (occupying the most grid cells)?

```{r lets.rangesize}
lets.rangesize(x = PAM, units = "cell") |>
  as.data.frame() |>
  arrange(desc(Range_size)) |>
  head(5)
```

What are the characteristics of the distribution range sizes?

```{r hist}
range_sizes <- lets.rangesize(x = PAM, units = "cell")

# Mean range size
mean(range_sizes)

# Proportion of species exceeding the mean range size
mean_range <- mean(range_sizes)
sum(range_sizes > mean_range) / nrow(range_sizes)

# Histogram of log-transformed range sizes
hist(
  log(range_sizes),
  main = "Distribution of species range sizes",
  xlab = "log(number of grid cells)"
)
```

Most species occupy relatively few grid cells; only a small fraction exceed the mean range size.

## Plot a Single Taxon Occurrence Map

```{r occ_map}
# Replace with any species name from PAM$Species_name
selected_taxon <- lets.rangesize(x = PAM, units = "cell") |>
  as.data.frame() |>
  arrange(desc(Range_size)) |>
  head(1)%>%row.names()
plot(PAM,
     name = selected_taxon,
     main = paste(selected_taxon, "occurrence"))
```

## Add Environmental Variables

Environmental variables can be integrated into the presence-absence matrix using `lets.addvar()`. The function handles spatial alignment automatically: cropping, aggregation, and extraction at grid cell centroids.

Here we use global annual mean temperature (°C) at 10-arcminute resolution from the *letsR* built-in dataset:

```{r PAM_env, message=FALSE, warning=FALSE}
temp <- terra::unwrap(letsR::temp)  # worldclim bio_1: annual mean temperature
PAM_env <- lets.addvar(PAM, temp, fun = mean)

# First 5 rows: longitude, latitude, and mean temperature
PAM_env[1:5, c(1, 2, ncol(PAM_env))]
```

Summarise mean temperature per species:

```{r temp_mean}
temp_mean <- lets.summarizer(PAM_env, ncol(PAM_env))

# Species with the warmest distributions
temp_mean |>
  arrange(desc(wc2.1_10m_bio_1_mean)) |>
  head(5)

# Species with the coldest distributions
temp_mean |>
  arrange(wc2.1_10m_bio_1_mean) |>
  head(5)
```

## Generate a Richness Map

```{r Richness_map}
plot(PAM,
     xlab = "Longitude",
     ylab = "Latitude",
     main = "Saxifraga richness map")
```

# References

Appelhans, Tim, Florian Detsch, Christoph Reudenbach, and Stefan Woellauer. 2023. “Mapview: Interactive Viewing of Spatial Data in r.” <https://CRAN.R-project.org/package=mapview>.

Boyle, Brad, Nicole Hopkins, Zhenyuan Lu, Juan Antonio Raygoza Garay, Dmitry Mozzherin, Tony Rees, Naim Matasci, et al. 2013. “The Taxonomic Name Resolution Service: An Online Tool for Automated Standardization of Plant Names.” BMC Bioinformatics 14 (1): 16. <https://doi.org/10.1186/1471-2105-14-16>.

Chirico, Michael. 2023. “geohashTools: Tools for Working with Geohashes.” <https://CRAN.R-project.org/package=geohashTools>.

Zizka, Alexander, Daniele Silvestro, Tobias Andermann, Josué Azevedo, Camila Duarte Ritter, Daniel Edler, Harith Farooq, et al. 2019. “CoordinateCleaner : Standardized Cleaning of Occurrence Records from Biological Collection Databases.” Edited by Tiago Quental. Methods in Ecology and Evolution 10 (5): 744–51. <https://doi.org/10.1111/2041-210X.13152>.

Vilela, Bruno, and Fabricio Villalobos. 2015. “letsR: A New r Package for Data Handling and Analysis in Macroecology.” *Methods in Ecology and Evolution* 6 (10): 1229–34. <https://doi.org/10.1111/2041-210x.12401>.
