---
title: "Example"
author:
  - name: Yuxuan Wang
    affiliation: Sichuan University
    email: wyx1743@gmail.com
    orcid: "0009-0003-7817-748X"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: false
    df_print: paged
    theme: readable
    highlight: tango
  rmarkdown::html_vignette:
    toc: false
  word_document: 
vignette: >
  %\VignetteIndexEntry{Example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

VasGBIF is a high-performance R package that compiles, validates, and standardizes tracheophyte occurrence records from GBIF into analysis-ready datasets. This basic example demonstrates the complete workflow using the built-in example dataset.

The VasGBIF pipeline consists of **4 stages** with **7 core functions**:

| Stage | Functions |
|------------------------------------|------------------------------------|
| **Stage 1:** Import and taxonomic standardisation | `import_records()`, `check_taxon()` |
| **Stage 2:** Duplicate detection via collection-event keys | `get_collections()`, `set_vouchers()` |
| **Stage 3:** Coordinate validation and native-status annotation | `refine_records()` |
| **Stage 4:** Export and visualisation | `export_records()`, `map_records()` |

After the three core stages, approximately 35% of the initial records are retained as high-quality, non-redundant data.

```{r init}
library(VasGBIF)
VasGBIF_summary <- list()
```

## Stage 1: Import and Taxonomic Standardisation

### Import Records

`import_records()` reads a GBIF occurrence download ZIP, extracts the occurrence table, and parses GBIF issue flags into record-level logical indicators.

The built-in example dataset is a GBIF download of *Medicago sativa* (preserved specimens with coordinates, obtained 23 July 2026): <https://doi.org/10.15468/dl.nt5exp>.

```{r import_records}
gbif_file <- system.file(
  "extdata", "0003386-260721160103020.zip",
  package = "VasGBIF"
)
occ_import <- import_records(path = gbif_file)
VasGBIF_summary$initial_records <- nrow(occ_import$occ)
VasGBIF_summary$initial_taxa <- occ_import$occ[, uniqueN(scientificName)]
```

The `import` object contains:

- `occ`: the occurrence records with core GBIF fields
- `occ_issue`: a binary matrix of GBIF issue flags per record
- `summary`: frequency table of detected issues
- `runtime`: execution time

### Check Taxon Name

`check_taxon()` resolves species- and infraspecific-rank names against the World Checklist of Vascular Plants (WCVP) via the Taxonomic Name Resolution Service (TNRS).

```{r check_taxon}
taxa_checked <- check_taxon(occ_import = occ_import, accuracy = 0.9)
VasGBIF_summary$taxa_submitted <- nrow(taxa_checked$occ_taxa_checked)
VasGBIF_summary$taxa_resolved <- taxa_checked$occ_taxa_checked[
  wcvp_searchNotes != "Not found", .N
]
```

The `occ_taxa` object contains:

- `occ_taxa_checked`: records with resolved WCVP taxonomy
- `summary`: resolution statistics by taxon rank
- `runtime`: execution time

## Stage 2: Duplicate Detection via Collection-Event Keys

### Generate Collection-Event Keys

`get_collections()` builds a composite key from the accepted taxon name, event date, and rounded coordinates (format: `wcvp_taxon_name|eventDate|latitude|longitude`). Records sharing the same key are treated as potential duplicates from a single gathering event.

```{r get_collections}
collection_keys <- get_collections(
  occ_import = occ_import,
  taxa_checked = taxa_checked,
  precision = 2L
)
VasGBIF_summary$complete_keys <- collection_keys$complete_keys
VasGBIF_summary$incomplete_keys <- collection_keys$incomplete_keys
```

The `collections` object contains:

- `occ_key`: the occurrence table with collection-event keys
- `complete_keys`: count of records with complete keys
- `incomplete_keys`: count of records with incomplete keys
- `runtime`: execution time

### Select Digital Vouchers

`set_vouchers()` scores each record on metadata completeness (9 fields) and geospatial quality (GBIF issue flags). The highest-scoring record in each duplicate group is designated the master digital voucher.

```{r set_vouchers}
voucher <- set_vouchers(
  occ_import = occ_import,
  taxa_checked = taxa_checked,
  collection_keys = collection_keys
)
VasGBIF_summary$usable <- voucher$occ_digital_voucher[
  VasGBIF_dataset_result == "usable", .N
]
VasGBIF_summary$duplicate <- voucher$occ_digital_voucher[
  VasGBIF_dataset_result == "duplicate", .N
]
```

The `vouchers` object contains:

- `occ_digital_voucher`: records with quality scores, grouping information, and `VasGBIF_dataset_result` (usable / duplicate)
- `occ_results`: quality assessment fields
- `runtime`: execution time

## Stage 3: Coordinate Validation and Native-Status Annotation

`refine_records()` performs three internal steps: restores missing metadata from duplicates, validates coordinates with CoordinateCleaner, and assigns native status by matching validated coordinates to WGSRPD Level 3 areas and WCVP distribution data.

```{r refine_records}
refined_records <- refine_records(voucher = voucher, threads = 4)
VasGBIF_summary$all_refined <- nrow(refined_records$all_records)
VasGBIF_summary$native <- refined_records$all_records[
  native_status == "native", .N
]
VasGBIF_summary$introduced <- refined_records$all_records[
  native_status == "introduced", .N
]
VasGBIF_summary$coordinate_problematic <- nrow(
  refined_records$CoordinateProblematic
)
VasGBIF_summary$final_taxa <- refined_records$all_records[
  , uniqueN(VasGBIF_wcvp_taxon_name)
]
```

The `refined` object contains:

- `all_records`: records with validated coordinates, restored metadata, and native status
- `CoordinateProblematic`: records that failed one or more CoordinateCleaner tests
- `runtime`: execution time

## Stage 4: Export and Visualisation

### Export Records

`export_records()` writes the refined records to disk as gzip-compressed CSV files.

```{r export_records, eval=FALSE, include=T}
export_records(
  refined_records = refined_records,
  export_path = getwd()
)
```

Three files are written:

- `usable_refined_records.csv.gz`: all records with validated coordinates
- `native_refined_records.csv.gz`: native subset
- `CoordinateProblematic_records.csv.gz`: records failing coordinate validation

### Map Records

`map_records()` renders an interactive map via `mapview`, with geohash-based decluttering and colour-coding by native status.

```{r map_records, eval=FALSE}
map_records(refined_records = refined_records, precision = 3, cex = 3)
```

## Summary

The summary statistics provide a quantitative overview of data reduction through the pipeline:

```{r summary}
VasGBIF_summary |>
  as.data.frame() |>
  t()
```

| Statistic | Description |
|----|----|
| `initial_records` | Total GBIF occurrence records imported |
| `initial_taxa` | Unique scientific names before resolution |
| `taxa_submitted` | Names submitted to TNRS |
| `taxa_resolved` | Names successfully resolved against WCVP |
| `complete_keys` | Records with complete collection-event keys |
| `incomplete_keys` | Records with incomplete keys |
| `usable` | Records classified as usable after voucher selection |
| `duplicate` | Records classified as duplicates |
| `all_refined` | Records retained after coordinate validation |
| `native` | Records classified as native |
| `introduced` | Records classified as introduced |
| `coordinate_problematic` | Records failing coordinate validation |
| `final_taxa` | Unique accepted taxon names in final dataset |

The workflow progressively filters the dataset through taxonomy resolution, duplicate grouping, quality scoring, and coordinate validation. The drop from `initial_records` to `all_refined` reflects the removal of duplicates and records with spatial issues. The transition from `initial_taxa` to `final_taxa` reflects synonym resolution and the removal of unresolvable names.

## Performance

VasGBIF achieves its speed through:

- **C/C++ backends**: core operations delegated to `data.table`, `stringi`, and `terra`
- **Vectorisation**: entire columns processed in single compiled calls via `fcase()`, `stri_detect_fixed()`
- **Memory-efficient design**: in-place modification (`:=`, `set()`) avoids unnecessary copies
- **Selective parallelisation**: CoordinateCleaner validation distributed across threads

On a standard laptop, VasGBIF compiles one million occurrence records within 15 minutes.

## References

GBIF.org (23 July 2026) GBIF Occurrence Download <https://doi.org/10.15468/dl.nt5exp>
