---
title: "Tutorial of UltraGBIF"
author:
  - name: anonymous
    affiliation: Anonymous University
    email: anonymous@anonymous.com
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Tutorial of UltraGBIF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

> **[IMPORTANT]**
>
> We have updated a more comprehensive tutorial, please visit <https://anonymous.4open.science/w/UltraGBIF-7FD3/>

## Overview

UltraGBIF is a high-performance R package designed to compile, validate, and standardize plant occurrence records from the Global Biodiversity Information Facility (GBIF) into analysis-ready datasets. This tutorial demonstrates the complete UltraGBIF workflow using [*Saxifraga*](https://doi.org/10.15468/dl.bythb4) occurrence records as an example dataset.

The UltraGBIF workflow consists of **4 stages** encompassing **7 functional modules**:

+-------------------------------------------------+-------------------------------------+-----------------------------+
| Stage                                           | Modules                             | Functions                   |
+=================================================+=====================================+=============================+
| **Stage 1:**                                    | 1.  Import Records                  | 1.  `import_records()`      |
|                                                 |                                     |                             |
| **Data Acquisition and Preparation**            | 2.  Check Taxon Name                | 2.  `check_occ_taxon()`     |
|                                                 |                                     |                             |
|                                                 | 3.  Check Collectors                | 3.  `check_collectors()`    |
+-------------------------------------------------+-------------------------------------+-----------------------------+
| **Stage 2:**                                    | 4.  Generate Unique Collection Mark | 4.  `set_collection_mark()` |
|                                                 |                                     |                             |
| **Duplicate Removal and Reliability Filtering** | 5.  Set Digital Voucher             | 5.  `set_digital_voucher()` |
+-------------------------------------------------+-------------------------------------+-----------------------------+
| **Stage 3:**                                    | 6.  Refine Records                  | 6.  `refine_records()`      |
|                                                 |                                     |                             |
| **Refine Records**                              |                                     |                             |
+-------------------------------------------------+-------------------------------------+-----------------------------+
| **Stage 4:**                                    | 7.  Map Records                     | 7.  `map_records()`         |
|                                                 |                                     |                             |
| **Visualization**                               |                                     |                             |
+-------------------------------------------------+-------------------------------------+-----------------------------+

After completing the three core stages (modules 1--6), approximately 30% of the initial occurrence records are typically retained as high-quality, non-redundant data.

## Stage 1: Data Acquisition and Preparation

This stage ensures data accuracy and consistency through three modules, laying the foundation for subsequent taxonomic and duplicate processing.

### Module 1: Import Records

This is the entry point of the UltraGBIF workflow. This module receives a path to the GBIF occurrence download zip file, loads all initial occurrence records, and automatically extracts GBIF issue flags for downstream quality assessment.

The function performs five sequential operations:

1.  **File validation**: Verifies that the input is a valid GBIF zip file
2.  **Decompression**: Automatically extracts the `occurrence.txt` file
3.  **Data loading**: Reads occurrence records using `data.table::fread()` with optimized column selection (55 core fields)
4.  **Issue extraction**: Parses the `issue` field to identify all GBIF quality flags
5.  **Summary compilation**: Generates a frequency table of all detected issues

The `issue` field in GBIF downloads contains a pipe-separated list of quality flags (e.g., `COORDINATE_ROUNDED|COUNTRY_COORDINATE_MISMATCH`). UltraGBIF transforms these into a structured binary matrix, enabling systematic quality assessment through individualized issue tracking and standardized classification.

```{r load, eval=FALSE, include=TRUE}
library(UltraGBIF)
library(curl,quietly = T)
library(data.table,quietly = T)
url <- 'https://api.gbif.org/v1/occurrence/download/request/0021523-250402121839773.zip'
gbif_occurrence_file <- paste0(path.expand("~"), '/Saxifraga.zip')
curl_download(url, gbif_occurrence_file, quiet = FALSE)
```

```{r import_records, eval=FALSE, include=TRUE}
message(gbif_occurrence_file)
UltraGBIF_summary <- list() ## save statistics information
occ_import <- import_records(
  GBIF_file = gbif_occurrence_file,
  only_PRESERVED_SPECIMEN = TRUE
)
UltraGBIF_summary$initial_records <- occ_import$occ[,.N]
UltraGBIF_summary$initial_taxa <- occ_import$occ[,uniqueN(Ctrl_scientificName)]
head(occ_import$summary,5)
```

The output `occ_import` is an `UltraGBIF_import` object containing:

-   `occ`: Initial occurrence records with 60 core fields (Ctrl-prefixed)
-   `occ_gbif_issue`: Structured binary matrix of GBIF issue flags
-   `summary`: Frequency table of all detected issues
-   `runtime`: Execution time

### Module 2: Check Taxon Name

Standardizes taxonomic names using the Taxonomic Name Resolution Service (TNRS) [@boyle2013] against the World Checklist of Vascular Plants (WCVP). The TNRS corrects spelling errors, resolves alternative spellings, and converts outdated synonyms to their current accepted names, enabling consistent taxonomic identification across the dataset.

The TNRS resolves names through four sequential steps:

1.  **Parse**: Tokenize the input name into genus and species epithet
2.  **Match**: Search for exact or fuzzy matches in the WCVP database
3.  **Correct**: Apply spelling corrections for fuzzy matches above the accuracy threshold
4.  **Select Best Match**: Choose the highest-scoring accepted name

```{r check_occ_taxon, eval=FALSE, include=TRUE}
taxa_checked <- check_occ_taxon(
  occ_import = occ_import,
  accuracy = 0.9
)
str(taxa_checked$summary[1])
```

The output `taxa_checked` is an `UltraGBIF_taxa_checked` object containing:

-   `occ_wcvp_check_name`: Records with resolved taxa, including accepted name, taxon rank, and WCVP ID
-   `summary`: Summary statistics of taxonomic resolution
-   `runtime`: Execution time

### Module 3: Check Collectors

Simplifies and extracts the primary collector's name from the `recordedBy` field, then builds a cleaned collector table. This module reduces inconsistencies that can fragment single collection events and improves the accuracy of subsequent duplicate detection.

The function processes the `recordedBy` field through the following steps:

1.  **Character encoding normalization**: Transliterates Cyrillic and other non-Latin scripts to Latin characters, then converts to ASCII
2.  **Special character removal**: Strips punctuation, symbols, and institutional keywords (e.g., "HERBARIUM", "EXPEDITION", "ET AL")
3.  **Primary name selection**: Extracts the longest token from the cleaned string, assuming it represents the primary collector's name
4.  **Threshold filtering**: Names shorter than `min_char` characters or matching a predefined exclusion list are classified as "UNKNOWN"

The function uses `tokenizers::tokenize_words()`[@tokenizers] for robust word tokenization. The longest word (rather than specifically the surname) is extracted as the primary collector identifier, filtering out initials, titles, and other short tokens.

```{r check_collectors, eval=FALSE, include=TRUE}
collectors_dictionary <- check_collectors(
  occ_import = occ_import,
  min_char = 2
)
UltraGBIF_summary$initial_collectors <- collectors_dictionary$collectors_dictionary[,.N]
UltraGBIF_summary$checked_collectors <- collectors_dictionary$collectors_dictionary[,uniqueN(Ctrl_nameRecordedBy_Standard)]
str(collectors_dictionary$collectors_dictionary)
```

The output `collectors_dictionary` is an `UltraGBIF_collectors_dictionary` object containing:

-   `collectors_dictionary`: Table mapping raw `recordedBy` strings to cleaned collector names
-   `runtime`: Execution time

## Stage 2: Duplicate Removal and Reliability Filtering

This stage improves data reliability by identifying high-quality, non-redundant occurrence records through two sequential modules.

### Module 4: Generate Unique Collection Mark

Generates a standardized composite key for each unique collection event. This module identifies and consolidates duplicates into unique collection events, where a collection event represents a distinct sampling instance (a specific collector at a specific date or with a specific record number).

The collection event key is constructed by concatenating three core metadata fields separated by underscores:

-   **Plant family** (`Ctrl_family`)
-   **Standardized collector name** (`Ctrl_nameRecordedBy_Standard`)
-   **Standardized collection number** (`Ctrl_recordNumber_Standard`)

The resulting identifier (`Ctrl_key`) follows the format `FAMILY_COLLECTORNAME_RECORDNUMBER`, grouping specimens sharing identical collection events.

Standardization ensures robust matching: non-numeric characters are stripped from collection numbers, and collector names are normalized using the standardized dictionary to reconcile spelling variants.

**Key Completeness Classification:**

Keys are classified based on the presence and validity of their components:

-   **Full keys**: Keys that do not end with an underscore and contain a valid collector name (i.e., not "UNKNOWN", "CAPTURED", or "ANONYMOUS")
-   **Incomplete keys**: Keys missing the collection number (ending with "\_") or containing unidentifiable collector names

**Fallback Strategy:**

When the collection number is unavailable but month information exists, the function attempts to use the event date (`Ctrl_eventDate`) as a surrogate identifier.

This module addresses two key challenges:

1.  Disambiguating common collector names through taxonomic context
2.  Enabling automated duplicate detection without reliance on often-missing temporal metadata

```{r set_collection_mark, eval=FALSE, include=TRUE}
collection_key <- set_collection_mark(
  occ_import = occ_import,
  collectors_dictionary = collectors_dictionary
)
UltraGBIF_summary$full_keys <- collection_key$full_keys[,.N]
UltraGBIF_summary$incomplete_keys <- collection_key$incomplete_keys[,.N]
```

The output `collection_key` is an `UltraGBIF_collection_key` object containing:

-   `collection_key`: Data table with `Ctrl_gbifID`, `Ctrl_nameRecordedBy_Standard`, `Ctrl_recordNumber_Standard`, and `Ctrl_key`
-   `full_keys`: Summary of complete keys with record counts, ordered by descending frequency
-   `incomplete_keys`: Summary of incomplete keys with record counts
-   `runtime`: Execution time

### Module 5: Set Digital Voucher

Resolves duplicate records and selects the master digital voucher to represent each collection event. For records possessing complete collection event keys (full keys), duplicates are identified and grouped; within each group, a master digital voucher is selected based on a composite quality score.

**Multi-Dimensional Quality Scoring System**

The quality scoring system evaluates each record across two dimensions:

1.  **Record Completeness**: Calculated as the sum of binary flags (1 = present, 0 = absent) for 10 metadata fields: `recordedBy`, `recordNumber`, `year`, `institutionCode`, `catalogNumber`, `locality`, `municipality`, `countryCode`, `stateProvince`, and `identifiedBy`.

2.  **Geospatial Quality**: Derived from GBIF geospatial issue flags (`EnumOccurrenceIssue`), classified into three severity levels:

    -   No Impact (`selection_score = -1`): Issues not affecting coordinate accuracy
    -   Potential Impact (`selection_score = -3`): Issues that may affect coordinate accuracy
    -   Exclusion (`selection_score = -9`): Records with severe geospatial issues

The final quality score combines these two dimensions, with the record exhibiting the highest score within each group selected as the digital voucher.

**Grouping Status Classification**

The `UltraGBIF_duplicates_grouping_status` field indicates the grouping outcome for each record:

1.  `groupable`: Complete collection key; duplicates identified and grouped
2.  `not groupable:`
    -   `no recordedBy and no recordNumber`: Both collector name and collection number are missing
    -   `no recordNumber`: Collection number is missing
    -   `no recordedBy`: Collector name is missing

Records with incomplete keys are treated as singleton events, with metadata completeness and geospatial scores computed identically to ensure consistent quality assessment.

**Coordinate Assignment**

For grouped records, if the master voucher has unusable coordinates (score = -9) or lacks coordinates entirely, usable coordinates from duplicates (score \>= -3) are preferentially retained and assigned to all members of the group.

**Taxonomic Resolution**

For groupable events, the accepted taxon name is assigned as the most frequently occurring species- or subspecies-level identification among duplicates; ties are resolved alphabetically. Unidentified events are flagged with `UltraGBIF_unidentified_sample = TRUE`.

```{r set_digital_voucher, eval=FALSE, include=TRUE}
voucher <- set_digital_voucher(
  occ_import = occ_import,
  taxa_checked = taxa_checked,
  collection_key = collection_key
)
UltraGBIF_summary$usable_records_before_refine <- voucher$occ_digital_voucher[UltraGBIF_dataset_result=='usable',.N]
gc()
```

The output `voucher` is an `UltraGBIF_voucher` object containing:

-   `occ_digital_voucher`: Full flagged occurrence records with quality scores, grouping information, and final classifications
-   `occ_results`: Quality assessment fields only
-   `runtime`: Execution time

Key output fields of `voucher$occ_digital_voucher` include:

-   `UltraGBIF_digital_voucher`: Logical; TRUE if the record is selected as the master voucher
-   `UltraGBIF_duplicates`: Logical; TRUE if the record belongs to a group with multiple members
-   `UltraGBIF_num_duplicates`: Integer; count of records within the same group
-   `UltraGBIF_duplicates_grouping_status`: Grouping feasibility classification
-   `UltraGBIF_dataset_result`: Final classification; one of "usable", "duplicate", or "unusable"

## Stage 3: Refine Records

This stage restores key information for usable records from their duplicates with the same collection mark, enhances geospatial accuracy, and extracts the native status information of occurrence records.

### Module 6: Refine Records

Validates spatial information, restores missing metadata from duplicates, and extracts native status information.

**Metadata Restoration**

For usable vouchers, missing information in key fields is restored by consolidating data from duplicate records belonging to identical collection events. This creates one clean record per unique event. The function selects the most common species-level name (using alphabetical order to break ties) or labels the event "unidentified" if no such name exists.

**Coordinate Validation**

Validated records are automatically subjected to geospatial validation using CoordinateCleaner [@zizka2019] and terra. This module efficiently performs coordinate validation and cleaning, targeting common geospatial anomalies:

-   Misassigned capitals
-   Geographic centroids
-   Duplicate coordinates
-   GBIF headquarters artifacts
-   Institution-proximate records
-   Distributional outliers
-   Marine points in terrestrial datasets
-   Zero/placeholder coordinates

**Native Status Extraction**

Each validated record is matched to its corresponding WGSRPD Level-3 Botanical Country classifications with different status (native, introduced, extinct, location_doubtful), minimizing mis-classification risks in cross-regional analyses.

**Parallel Processing**

Coordinate validation is distributed across multiple threads using `foreach`/`doParallel` to accelerate processing of large datasets.

```{r refine_records, eval=FALSE, include=TRUE}
refined_records <- refine_records(
  voucher = voucher,
  threads = 4,
  save_path = dirname(gbif_occurrence_file),
  tests = c("capitals", "centroids", "equal", "gbif",
            "institutions", "outliers", "seas", "zeros")
)
UltraGBIF_summary$usable_records_after_refine <- refined_records$native_records[,.N]+refined_records$other_records[,.N]
UltraGBIF_summary$final_taxa <- uniqueN(c(refined_records$native_records[,UltraGBIF_wcvp_taxon_name],refined_records$other_records[,UltraGBIF_wcvp_taxon_name]))
```

The output `refined_records` is an `UltraGBIF_refined` object containing:

-   `native_records`: Records classified as native to their geographic region
-   `other_records`: Records classified as introduced, extinct, location_doubtful, or unknown
-   `runtime`: Execution time

When `save_path` is specified, `native_records` and `other_records` are exported as compressed CSV files (`native_refined_records.csv.gz` and `usable_refined_records.csv.gz`).

### Summary core modules

```{r summary, eval=FALSE, include=TRUE}
library(dplyr)
UltraGBIF_summary%>%as.data.frame()%>%t()
```

The summary statistics provide a quantitative overview of the data reduction throughout the UltraGBIF workflow:

+----------------------------------+--------------------------------------------------------------------+
| Statistic                        | Description                                                        |
+==================================+====================================================================+
| `initial_records`                | Total number of GBIF occurrence records imported initially         |
+----------------------------------+--------------------------------------------------------------------+
| `initial_taxa`                   | Number of unique scientific names before taxonomy resolution       |
+----------------------------------+--------------------------------------------------------------------+
| `initial_collectors`             | Number of unique collector names before standardization            |
+----------------------------------+--------------------------------------------------------------------+
| `checked_collectors`             | Number of unique collector names after standardization             |
+----------------------------------+--------------------------------------------------------------------+
| `full_keys`                      | Number of complete collection event keys (groupable)               |
+----------------------------------+--------------------------------------------------------------------+
| `incomplete_keys`                | Number of incomplete keys (non-groupable)                          |
+----------------------------------+--------------------------------------------------------------------+
| `usable_records_before_refine`   | Number of records classified as usable after voucher selection     |
+----------------------------------+--------------------------------------------------------------------+
| `usable_records_after_refine`    | Number of records retained after refinement                        |
+----------------------------------+--------------------------------------------------------------------+
| `final_taxa`                     | Number of unique accepted taxon names in the final refined dataset |
+----------------------------------+--------------------------------------------------------------------+

The workflow progressively filters the initial dataset through taxonomy resolution, collector name normalization, duplicate grouping, and quality scoring. The transition from `initial_taxa` to `final_taxa` reflects synonym resolution and the removal of unresolvable or invalid names. Similarly, the reduction from `initial_collectors` to `checked_collectors` demonstrates the consolidation of spelling variants and formatting inconsistencies in collector metadata. The ratio of `full_keys` to `incomplete_keys` indicates the proportion of records with sufficient information for reliable duplicate detection. After completing modules 1\~6, the core UltraGBIF workflow is fully executed, yielding a high-quality, non-redundant dataset ready for downstream analysis.

## Stage 4: Visualization

Two optional modules are available for data exploration and visualization after completing the core pipeline.

### Module 7: Map Records

Renders verified records onto customizable, dynamic interactive maps using `mapview` [@mapview]. This module employs `geohashTools` [@geohashTools] to perform spatial clustering of geo-coordinates.

The `precision` parameter determines grid resolution e.g.:

-   Precision 4: approximately 20 km x 20 km grids
-   Precision 3: approximately 156 km x 156 km grids

Records are grouped by taxon name, WGSRPD information, and geohash grid, with deduplication applied. Dynamic maps support:

-   Multiple basemap layers (OpenStreetMap, Esri World Imagery, Stadia Stamen Watercolor)
-   WGSRPD information legend
-   Adjustable symbol size via the `cex` parameter
-   Zoom functionality for spatial resolution adjustment
-   Clickable popups with record metadata

```{r map_records, eval=FALSE, include=TRUE}
map_records(
  refined_records = refined_records,
  precision = 2,
  cex = 4
)
```

## Minimal Complete Workflow

The following code demonstrates the complete UltraGBIF workflow from data import to richness mapping:

```{r complete_workflow, eval=FALSE, include=TRUE}
# Step 1: Import GBIF occurrence records
occ_import <- import_records(
  GBIF_file = "path/to/gbif_download.zip",
  only_PRESERVED_SPECIMEN = TRUE
)

# Step 2: Standardize taxonomic names
taxa_checked <- check_occ_taxon(
  occ_import = occ_import,
  accuracy = 0.9
)

# Step 3: Standardize collector names
collectors_dictionary <- check_collectors(
  occ_import = occ_import,
  min_char = 2
)

# Step 4: Generate collection event keys
collection_key <- set_collection_mark(
  occ_import = occ_import,
  collectors_dictionary = collectors_dictionary
)

# Step 5: Select digital vouchers
voucher <- set_digital_voucher(
  occ_import = occ_import,
  taxa_checked = taxa_checked,
  collection_key = collection_key
)

# Step 6: Refine records (coordinate validation + native status)
refined_records <- refine_records(
  voucher = voucher,
  threads = 4,
  save_path = getwd()
)

# Optional: Visualize results
map_records(refined_records = refined_records, precision = 4, cex = 4)

```

## Performance

UltraGBIF achieves outstanding performance through specific technical architectures:

-   **C/C++ Backend Integration**: Leverages `data.table`, `stringi`, and `terra`, all implemented in C/C++
-   **Vectorization Over Explicit Loops**: Bypasses R's interpretive overhead by dispatching to pre-compiled routines
-   **SIMD Exploitation**: Vectorized operations enable compiler-level SIMD auto-vectorization
-   **Memory-Efficient Design**: In-place modification (`set()`, `:=`) eliminates intermediate copies
-   **Chunk-Based Parallelization**: Vectorized processing within chunks, parallel execution across chunks

On a standard laptop, UltraGBIF can compile one million occurrence records within 15 minutes.

## References
